Short-term synaptic and neuronal plasticity act in synergy for deviance detection in spiking networks

### 2022-12-14 paper-1/manuscript.tex
Editing

To investigate deviance detection, we used a classical oddball paradigm as illustrated in Figure \ref{fig1}d, presenting a total of 500 stimuli at regular intervals of 500~ms. Target stimuli (labeled A hereafter) and distractor stimuli (labeled B to E) were presented in three randomized sequences: The "standard" sequence (std), consisting of 400 presentations of A and 100 presentations of B; the "deviant" sequence (dev) with 100 A and 400 B, and the many-standards control sequence (msc) with 100 presentations of each of the stimuli A through E. In this way, A is presented as the predominant and therefore expected stimulus (std), as an infrequent violation of an expectation (of B, dev), and, to control for pure adaptation effects, as an equally infrequent stimulus with no strong expectation of any particular input (msc).

@@

\begin{figure*}%
    \centering
    \includegraphics{fig4.pdf}
    \caption{}
    \label{fig4}
\end{figure*}
\textbf{Fig. 4} Responses and threshold adaptation with TA only.
\textbf{a} Target trial average post-stimulus spike histogram, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and conditions. Left and middle column, target trials in dev sequence and msc sequence, respectively; right column, contrast between these two.
\textbf{b} Target trial average of the threshold adaptation voltage $V_{TA}$, with axes, neuron order and columns as in panel a.
\textbf{c} Target trial average of the membrane potential $V_m$, with axes, neuron order and columns as in panel a. The first 10 ms of the neurons under direct stimulation are masked out to avoid saturation of the color mask.
\textbf{d} Relationship across excitatory neurons between the contrast (dev - msc) in $V_{TA}$ at the start of target trials and the contrast in average number of spikes fired in target trials, and associated linear regression.
\textbf{e} Mean contrast (dev - msc) in $V_{TA}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\\Delta V_{TA}$ vs response size contrast across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant correlation (p < 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
\textbf{f} In the sample network, contribution of the contrast in $V_{TA}$ and $V_m$ to bins with increased firing in dev (i.e., positive/red bins in the spike probability contrast, panel a), weighted by that same contrast.
\textbf{g} Contribution of $V_{TA}$ relative to the sum of the absolute values of the weighted contributions in panel f.
\textbf{h} Relative contribution of $V_{TA}$ across networks and stimuli, showing median (solid line) and inter-quartile range (shaded area)

@@

\begin{figure*}%
    \centering
    \includegraphics{fig5.pdf}
    \caption{}
    \label{fig5}
\end{figure*}
\textbf{Fig. 5} Non-target response magnitude determines thresholds.
\textbf{a} Sample network $V_{TA}$ at the start of A trials (mean over trials), with each neuron shown in its spatial location. Left and middle column, raw averages for dev and msc sequences, respectively; right column, contrast between dev and msc averages. A and B stimulus locations for this and following figures are highlighted in the left plot; see Figure \ref{fig1} for the locations of the remaining stimuli used in the msc sequence.
\textbf{b} Average response (in spikes per trial) to non-target trials (B in dev, B, C, D, and E in msc). Columns are arranged as in panel a.
\textbf{c} Relationship across excitatory neurons between the contrast (dev - msc) in average number of spikes fired in response to non-target stimulation and the contrast in $V_{TA}$ at the start of target trials, and associated linear regression.
\textbf{d} Mean contrast (dev - msc) in the non-target response magnitude across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel (non-target response size contrast vs $\\Delta V_{TA}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant correlation (p < 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations

\begin{figure*}%
    \centering
    \includegraphics{fig6.pdf}
    \caption{}
    \label{fig6}
\end{figure*}
\textbf{Fig. 6} Stimulus replacement and adaptation both lead to reduced dev non-target response.
\textbf{a} Non-target response contrasts: The replacement contrast (top right) compares the response to B in msc (top left) to all non-target responses in msc (top center). The adaptation contrast (bottom right) compares the response to B in dev (bottom left) to the response to B in msc (bottom center). Both contrasts together add up to the non-target contrast (dev - msc) shown in Figure \ref{fig5}b. Note that the contrast column (right) shares a color scale to allow a direct comparison.
\textbf{b} Mean replacement (left) and adaptation (right) contrasts across networks and stimuli, as well as the mean replacement contrast filtered for neurons near B, i.e., neurons within one excitatory synapse of the neurons under direct stimulation of B (center).
\textbf{c} Mean $V_{TA}$ contrast (dev - msc) at the start of A trials, filtered for neurons near B. Each marker corresponds to one network; the sample network is highlighted in red. The histogram in the background shows the distribution of the data, with the orange portion indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations

>>>>

To investigate deviance detection, we used a classical oddball paradigm as illustrated in Figure \ref{fig1}d, presenting a total of 500 stimuli at regular intervals of 500~ms. Target stimuli (labeled "A" or "target" hereafter) and distractor stimuli (labeled B to E) were presented in three randomized sequences: The "standard" sequence (std), consisting of 400 presentations of A and 100 presentations of B; the "deviant" sequence (dev) with 100 A and 400 B, and the many-standards control sequence (msc) with 100 presentations of each of the stimuli A through E. In this way, A is presented as the predominant and therefore expected stimulus (std), as an infrequent violation of an expectation (of B, dev), and, to control for pure adaptation effects, as an equally infrequent stimulus with no strong expectation of any particular input (msc).

@@

\begin{figure*}%
    \centering
    \includegraphics{fig4.pdf}
    \caption{}
    \label{fig4}
\end{figure*}
\textbf{Fig. 4} Responses and threshold adaptation with TA only.
\textbf{a} Target trial average post-stimulus spike histogram in the sample network, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and conditions. Left and middle column, target trials in dev sequence and msc sequence, respectively; right column, contrast between these two.
\textbf{b} Target trial average of the threshold adaptation voltage $V_{TA}$, with axes, neuron order and columns as in panel a.
\textbf{c} Target trial average of the membrane potential $V_m$, with axes, neuron order and columns as in panel a. The first 10 ms of the neurons under direct stimulation are masked out to avoid saturation of the color mask.
\textbf{d} Relationship across excitatory neurons of the sample network between the contrast (dev - msc) in $V_{TA}$ at the start of target trials and the contrast in average number of spikes fired in target trials, and associated linear regression.
\textbf{e} Network-wide mean contrast (dev - msc) in $V_{TA}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\\Delta V_{TA}$ vs response size contrast across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant correlation (p < 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
\textbf{f} In the sample network, contribution of the contrast in $V_{TA}$ and $V_m$ to bins with increased firing in dev (i.e., positive/red bins in the spike probability contrast, panel a), weighted by that same contrast.
\textbf{g} Contribution of $V_{TA}$ relative to the sum of the absolute values of the weighted contributions in panel f.
\textbf{h} Relative contribution of $V_{TA}$ across networks and stimuli, showing median (solid line) and inter-quartile range (shaded area)

@@

\begin{figure*}%
    \centering
    \includegraphics{fig5.pdf}
    \caption{}
    \label{fig5}
\end{figure*}
\textbf{Fig. 5} Non-target response magnitude determines thresholds.
\textbf{a} $V_{TA}$ at the start of A trials (mean over trials) in the sample network, with each neuron shown in its spatial location. Left and middle column, raw averages for dev and msc sequences, respectively; right column, contrast between dev and msc averages. A and B stimulus locations for this and following figures are highlighted in the left plot; see Figure \ref{fig1} for the locations of the remaining stimuli used in the msc sequence.
\textbf{b} Average response (in spikes per trial) to non-target trials (B in dev, B, C, D, and E in msc). Columns are arranged as in panel a.
\textbf{c} Relationship across excitatory neurons of the sample network between the contrast (dev - msc) in average number of spikes fired in response to non-target stimulation and the contrast in $V_{TA}$ at the start of target trials, and associated linear regression.
\textbf{d} Network-wide mean contrast (dev - msc) in the non-target response magnitude across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel (non-target response size contrast vs $\\Delta V_{TA}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant correlation (p < 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations

\begin{figure*}%
    \centering
    \includegraphics{fig6.pdf}
    \caption{}
    \label{fig6}
\end{figure*}
\textbf{Fig. 6} Stimulus replacement and adaptation both lead to a reduced dev non-target response.
\textbf{a} Non-target response contrasts in the sample network: The replacement contrast (top right) compares the response to B in msc (top left) to all non-target responses in msc (top center). The adaptation contrast (bottom right) compares the response to B in dev (bottom left) to the response to B in msc (bottom center). Both contrasts together add up to the non-target contrast (dev - msc) shown in Figure \ref{fig5}b. Note that the contrast column (right) shares a color scale to allow a direct comparison.
\textbf{b} Network-wide mean replacement (left) and adaptation (right) contrasts across networks and stimuli, as well as the mean replacement contrast filtered for neurons near B, i.e., neurons within one excitatory synapse of the neurons under direct stimulation of B (center).
\textbf{c} Mean $V_{TA}$ contrast (dev - msc) at the start of A trials, filtered for neurons near B. Each marker corresponds to one network; the sample network is highlighted in red. The histogram in the background shows the distribution of the data, with the orange portion indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations
